---
title: "Gas Prices"
---

```{r, include=FALSE, eval=TRUE, warning=FALSE}
# Gas prices
df.rate <- df.miles

# Read file from eia.gov
df.gas <- read.csv("../data/U.S._All_Grades_All_Formulations_Retail_Gasoline_Prices_Monthly_Dec2021.csv",
                   skip = 5, header=FALSE) %>%
  mutate(date=my(V1)) %>% rename(mrate=V2) %>% select(date,mrate) %>%
  filter(date >= min(df.rate$date))

# Combine
df.rate <- df.rate %>% left_join(df.gas)

#
# Plot gas prices over time
#
p.rate <- ggplot(df.rate) +
  geom_segment(aes(x=date, xend=date, y=rate, yend=mrate), size=0.3,color="grey75",alpha=0.7) +
  geom_line(aes(x=date, y=mrate), size=1.2, color="grey75", alpha=0.7) +
  geom_point(aes(x=date, y=rate), color="grey25", size=1.0, alpha=0.8) +
  #geom_line(aes(x=date, y=cor), color="indianred", size=1.2, alpha=0.8) +
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Date", y="Gas Price") +
  theme_minimal()

```

The plot below compares what we paid for gas with the national average monthly gas price according to the [US Energy Information Administration](https://www.eia.gov/opendata/qb.php?sdid=PET.EMM_EPM0_PTE_NUS_DPG.M).  Each data point represents the monthly aggregate for all gas purchased during a particular month, and the trendline indicates the national average.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
p.rate
```

<details>
<summary> <b> Details: Seasonal Pattern </b> </summary>

```{r, include=FALSE}

# Create a timeseries object
ts.rate <- na.interp(ts(df.rate$rate, start = c(2005,4),frequency=12))
fc.rate <- decompose(ts.rate)
df.rate$trend <- as.numeric(fc.rate$trend)
df.rate$seasonal <- as.numeric(fc.rate$seasonal)
df.rate$cor <- as.numeric(fc.rate$x) - as.numeric(fc.rate$seasonal)

# 
df.rt <- df.rate %>% group_by(month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal, na.rm=TRUE)) %>%
  mutate(month = as.factor(month))

p.srate <- ggplot(df.rt) +
  geom_point(aes(x=month, y=seasonal), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=1), size=1, alpha=0.2) + 
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Month", y="Seasonal change in gas price") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))

pal.smiles <- c("TRUE"="#96aff2", "FALSE"="#f8988f")
p.srate2 <- ggplot(df.rt) +
  geom_col(aes(x=month, y=seasonal, fill=seasonal>=0)) +
  scale_y_continuous(labels=dollar_format()) +
  scale_fill_manual(values=pal.smiles) +
  labs(x="Month", y="Seasonal change in gas price") +
  theme_minimal() +
  theme(axis.text = element_text(size=10), 
        legend.position="none",
        strip.text = element_blank())
```

<br />
The seasonal fluctuations in the gas prices amount to <b> `r dollar(max(df.rt$seasonal)-min(df.rt$seasonal)) ` </b>, with the peak price in <b> `r month(which.max(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>, and the lowest in <b> `r month(which.min(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>. 

```{r, echo=FALSE}
p.srate2
```
</details>

<br>
