---
title: "Gas Mileage"
---

```{r, include=FALSE, message=FALSE, echo=FALSE}
#
# NOT INCLUDED
# time series analysis of total miles
#
source("_site.R")
# Create a timeseries object
first.year <- year(min(df.miles$date))
first.month <- month(min(df.miles$date))

# Create a time series object
ts.miles <- na.interp(ts(df.miles$miles, start = c(first.year,first.month),frequency=12))
dcmp.a <- decompose(ts.miles, "additive")
dcmp.m <- decompose(ts.miles, "multiplicative")

df.miles$trend <- dcmp.a$trend

p.mileage <- ggplot(df.miles) +
  geom_point(aes(x=date, y=miles), color="grey75", size=2, alpha=0.6) +
  geom_line(aes(x=date, y=trend), color="indianred", size=2, alpha=0.8) +
  scale_y_continuous(limits=c(0,1.1*max(df.miles$miles))) +
  labs(x="Date", y="Monthly Miles") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

```{r, message=FALSE, include=FALSE, warning=FALSE}
p.mileage
```


The plot below shows the monthly gas mileage for each vehicle over this span.  Each data point represents the monthly aggregate for a particular vehicle, and the curves are smoothed lines obtained from a time-series decomposition.

```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
# Aggregate by date & car, get mean mileage & gas price
df.grp <- df.cars %>% 
  group_by(date = floor_date(ymd(rmT(Date)), unit="month"), car) %>%
  summarise(mileage = sum(miles)/sum(gal),
            rate = mean(rate), miles=sum(miles), cost=sum(cost)) 


# Fill in missing months
df.cv <- get_missingmonths(df.grp[ df.grp$car == "cav", ]) %>% mutate(car = "cav")
df.cr <- get_missingmonths(df.grp[ df.grp$car == "cor", ]) %>% mutate(car = "cor")
df.rv <- get_missingmonths(df.grp[ df.grp$car == "rav", ]) %>% mutate(car = "rav")

# Create a timeseries objects
ts.cav <- na.interp(ts(df.cv$mileage[ df.cv$car=="cav" ], 
                       start = c(2005,4),frequency=12))
ts.cor <- na.interp(ts(df.cr$mileage[ df.cr$car=="cor" ], 
                       start = c(2008,11),frequency=12))
ts.rav <- na.interp(ts(df.rv$mileage[ df.rv$car=="rav" ], 
                       start = c(2018,7),frequency=12))

# Time series decomposition
fc.cav <- decompose(ts.cav, type="additive")
fc.cor <- decompose(ts.cor, type="additive")
fc.rav <- decompose(ts.rav, type="additive")

# Add time series results to dataframes
df.cv <- df.cv %>% mutate(trend = as.numeric(fc.cav$trend),
                          seasonal = as.numeric(fc.cav$seasonal),
                          cor = as.numeric(fc.cav$x) - seasonal)
df.cr <- df.cr %>% mutate(trend = as.numeric(fc.cor$trend),
                          seasonal = as.numeric(fc.cor$seasonal),
                          cor = as.numeric(fc.cor$x) - seasonal)
df.rv <- df.rv %>% mutate(trend = as.numeric(fc.rav$trend),
                          seasonal = as.numeric(fc.rav$seasonal),
                          cor = as.numeric(fc.rav$x) - seasonal)

# Merge it all together
df.mileage <- rbind(df.cv, df.cr, df.rv)

#
# Plot monthly gas mileage for each vehicle 
#
p.mlg2 <- ggplot(df.mileage) + 
  geom_point(aes(x=date, y=mileage, group=car, color=car), size=1.2, alpha=0.5) + 
  geom_line(aes(x=date, y=trend, group=car, color=car), size=1.2, alpha=0.8) + 
  annotate("text",x=as.Date("2006-02-01"),y=34,label="Cavalier",
           color="#f8988f",size=5,alpha=0.9) +
  annotate("text",x=as.Date("2010-01-01"),y=37,label="Corolla",
         color="#74c973",size=5,alpha=0.9) +
  annotate("text",x=as.Date("2020-01-01"),y=18,label="Rav4",
         color="#96aff2",size=5,alpha=0.9) +
  scale_y_continuous(limits = c(15,40)) +
  labs(x="Date",y="Gas Mileage (mpg)") +
  theme_minimal() +
  theme(legend.position="none") +
  theme(axis.text = element_text(size=10))
```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
p.mlg2
```

<details>
<summary> <b> Details: Gas Mileage </b> </summary>
<br />
These histograms compare the monthly gas mileage aggregates to values quoted on https://www.fueleconomy.gov/ for each vehicle.

```{r, include=FALSE}

df.fuel <- data.frame(car = c("cav","cor","rav"),
                      city = c(21,27,22),
                      hwy = c(29,35,28),
                      cmb = c(24,30,25),
                      lab = c("Cavalier","Corolla","RAV4"))

p.mlg <- ggplot(df.cars) +
  geom_segment(data=df.fuel, aes(x=city, xend=city, y=0, yend=50, color=car)) +
  geom_text(data=df.fuel, aes(x=city, y=45, label="city",color=car), hjust=1.2,alpha=0.8) +
  geom_text(data=df.fuel, aes(x=hwy, y=45, label="hwy",color=car), hjust=1.2,alpha=0.8) +
  geom_text(data=df.fuel, aes(x=40, y=40, label=lab,color=car), hjust=1.2, size=5) +
  geom_segment(data=df.fuel, aes(x=hwy, xend=hwy, y=0, yend=50, color=car)) +
  geom_histogram( aes(x=mileage, fill=car), bins=40, alpha=0.8) +
  labs(x="Gas Mileage (mpg)", y="Count") +
  facet_wrap( ~car, ncol=1) +
  theme_minimal() +
  theme(legend.position="none", strip.text = element_blank())
  
```

```{r, echo=FALSE}
p.mlg
```


</details>

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE}

# Time series object was created above
# Seasonal component to gas mileage
df.smileage <- df.mileage %>% 
  group_by(car, month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal))

p.smileage <- ggplot(df.smileage) + 
  geom_point(aes(x=month, y=seasonal, group=car, color=car), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=car, color=car), size=1, alpha=0.2) + 
  #scale_y_continuous(limits = c(15,40)) +
  labs(x="Month",y="Seasonal change in gas mileage (mpg)") +
  theme_minimal() +
  theme(legend.position="none",axis.text = element_text(size=10))

diffcav <- max(df.smileage$seasonal[df.smileage$car == "cav"]) - 
           min(df.smileage$seasonal[df.smileage$car == "cav"])
diffcor <- max(df.smileage$seasonal[df.smileage$car == "cor"]) - 
           min(df.smileage$seasonal[df.smileage$car == "cor"])
diffrav <- max(df.smileage$seasonal[df.smileage$car == "rav"]) - 
           min(df.smileage$seasonal[df.smileage$car == "rav"])

```

<br />
Seasonal fluctuations in gas mileage were determined from a time series decomposition.  The values vary for each vehicle:  
- For the <b>Cavalier</b> the magnitude is  <b>`r format(diffcav,digits=1,nsmall=1)` mpg</b>  
- For the <b>Corolla</b> the magnitude is <b>`r format(diffcor,digits=1,nsmall=1)` mpg</b>   
- For the <b>RAV4</b> the magnitude is <b>`r format(diffrav,digits=1,nsmall=1)` mpg</b>   

```{r, echo=FALSE}
p.smileage
```
</details>

<br />