---
title: "Analysis of Gas Receipts"
author: "Nathan Hoteling"
date: "`r format(Sys.time(), '%B %Y')`"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(ggplot2)
library(dplyr)
library(lubridate)
library(reshape2)
library(scales)
library(plotly)

library(forecast)

GAS_PRICES <- "../data/U.S._Regular_All_Formulations_Retail_Gasoline_Prices_Monthly_Apr2023.csv"
```

```{r, include=FALSE}
df.cav <- read.table("../data/cavalier_2001.csv",sep=",",header=TRUE,fill=TRUE)
df.cor <- read.table("../data/corolla_2009.csv",sep=",",header=TRUE,fill=TRUE)
df.rav <- read.table("../data/rav4_2018.csv",sep=",",header=TRUE,fill=TRUE)
df.hdr <- read.table("../data/highlander_2023.csv",sep=",",header=TRUE,fill=TRUE)

df.cav$car <- "cav"
df.cor$car <- "cor"
df.rav$car <- "rav"
df.hdr$car <- "hdr"

df.cars <- dplyr::bind_rows(df.cav, df.cor, df.rav, df.hdr)

# Function to remove "T" from date string
# --> date strings are appended with "T" to make them Excel-safe
rmT <- function(fstr) { stringr::str_replace(fstr, "T", "") }

# Function to fill in monthly data where no monthly data exist
get_missingmonths <- function(df) {
  dfx <- data.frame(date = seq.Date(from = min(df$date, na.rm=TRUE), 
               to = max(df$date, na.rm=TRUE), 
               by="month")) %>%
    left_join(df, by="date")
  return(dfx)
}
```


```{r, include=FALSE}
# aggregate data by month
df.bymonth <- df.cars %>% 
  group_by(date = floor_date(ymd(rmT(Date)), unit="month")) %>%
  summarise(miles = sum(miles), cost=sum(cost), rate = mean(rate), gal=sum(gal)) 


# Fill in gaps
df.miles <- get_missingmonths(df.bymonth) %>%
  mutate(label = paste0("Date: ", format.Date(date, "%Y-%m"),"<br>",
                        "Miles: ", format(miles, digits=1, big.mark=",")))

p.miles <- ggplot(df.miles, aes(text=label, group=1)) +
  geom_line(aes(x=date, y=miles), color="grey20") +
  scale_y_continuous(limits=c(0,1.1*max(df.miles$miles)), labels=comma) +
  labs(x="Date", y="Monthly Miles") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))

# interval, for use below
t <- interval(min(df.bymonth$date, na.rm=TRUE), max(df.bymonth$date, na.rm=TRUE))
tlen <- time_length(t, unit="year")
avg_miles <- sum(df.bymonth$miles)/tlen
avg_cost <- sum(df.bymonth$cost)/tlen
tot_miles <- sum(df.bymonth$miles)
tot_cost <- sum(df.bymonth$cost)
```

<hr>
<br>

This analysis provides a summary of data collected from gas receipts for all vehicles I have owned from <b> `r format(min(df.bymonth$date, na.rm=TRUE), "%B %Y")` </b> to <b> `r format(max(df.bymonth$date, na.rm=TRUE), "%B %Y")`</b>, a span of <b> `r format(tlen, digits=1,nsmall=1)` </b> years.  Over this time period we drove a total of <b> `r format(tot_miles, digits=0,nsmal=0, scientific=FALSE, big.mark=",")` </b> miles and spent <b> `r dollar(tot_cost)` </b> on gas.

<br>
<br>

## Total Miles

The plot below shows the number of miles driven per month for all vehicles over this period.  Note that gaps in the line chart represent months with no data; in some cases this was because we didn't purchase gas during that month and in others it is because gas receipts were lost.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
ggplotly(p.miles, tooltip="text")
```

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE, message=FALSE, warning=FALSE}

# Create a timeseries object
ts.miles <- na.interp(ts(df.miles$miles, start = c(2005,4),frequency=12))
fc.miles <- decompose(ts.miles)
df.miles$trend <- as.numeric(fc.miles$trend)
df.miles$seasonal <- as.numeric(fc.miles$seasonal)
df.miles$cor <- as.numeric(fc.miles$x) - as.numeric(fc.miles$seasonal)

# 
df.smiles <- df.miles %>% group_by(month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal, na.rm=TRUE)) %>%
  mutate(month = as.factor(month))

p.smiles <- ggplot(df.smiles) +
  geom_point(aes(x=month, y=seasonal), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=1), size=1, alpha=0.2) + 
  labs(x="Month", y="Seasonal change in miles driven") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))

pal.smiles <- c("TRUE"="#96aff2", "FALSE"="#f8988f")
p.smiles2 <- ggplot(df.smiles) +
  geom_col(aes(x=month, y=seasonal, fill=seasonal>=0)) +
  scale_fill_manual(values=pal.smiles) +
  labs(x="Month", y="Seasonal change in miles driven") +
  theme_minimal() +
  theme(axis.text = element_text(size=10), legend.position="none")
```

<br>

There are some seasonal patterns in the number of miles driven over the course of the year. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
#ggplotly(p.smiles2)
p.smiles2
```
</details>

<br>
<br>
<br>

## Gas Mileage

```{r, include=FALSE, message=FALSE, echo=FALSE, warning=FALSE}
#
# NOT INCLUDED
# time series analysis of total miles
#

# Create a timeseries object
first.year <- year(min(df.miles$date))
first.month <- month(min(df.miles$date))

# Create a time series object
ts.miles <- na.interp(ts(df.miles$miles, start = c(first.year,first.month),frequency=12))
dcmp.a <- decompose(ts.miles, "additive")
dcmp.m <- decompose(ts.miles, "multiplicative")

df.miles$trend <- dcmp.a$trend

p.mileage <- ggplot(df.miles) +
  geom_point(aes(x=date, y=miles), color="grey75", size=2, alpha=0.6) +
  geom_line(aes(x=date, y=trend), color="indianred", size=2, alpha=0.8) +
  scale_y_continuous(limits=c(0,1.1*max(df.miles$miles))) +
  labs(x="Date", y="Monthly Miles") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

```{r, message=FALSE, include=FALSE, warning=FALSE}
p.mileage
```


The plot below shows the monthly gas mileage for each vehicle over this span.  Each data point represents the monthly aggregate for a particular vehicle, and the curves are smoothed lines obtained from a time-series decomposition.

```{r, echo=FALSE, include=FALSE, message=FALSE, warning=FALSE}
# Aggregate by date & car, get mean mileage & gas price
df.grp <- df.cars %>% 
  group_by(date = floor_date(ymd(rmT(Date)), unit="month"), car) %>%
  summarise(mileage = sum(miles)/sum(gal),
            rate = mean(rate), miles=sum(miles), cost=sum(cost)) 


# Fill in missing months
df.cv <- get_missingmonths(df.grp[ df.grp$car == "cav", ]) %>% mutate(car = "cav")
df.cr <- get_missingmonths(df.grp[ df.grp$car == "cor", ]) %>% mutate(car = "cor")
df.rv <- get_missingmonths(df.grp[ df.grp$car == "rav", ]) %>% mutate(car = "rav")
df.hd <- get_missingmonths(df.grp[ df.grp$car == "hdr", ]) %>% mutate(car = "hdr")

# Create a timeseries objects
ts.cav <- na.interp(ts(df.cv$mileage[ df.cv$car=="cav" ], 
                       start = c(2005,4),frequency=12))
ts.cor <- na.interp(ts(df.cr$mileage[ df.cr$car=="cor" ], 
                       start = c(2008,11),frequency=12))
ts.rav <- na.interp(ts(df.rv$mileage[ df.rv$car=="rav" ], 
                       start = c(2018,7),frequency=12))
# ts.hdr <- na.interp(ts(df.hd$mileage[ df.hd$car=="hdr" ], 
#                        start = c(2023,3),frequency=12))

# Time series decomposition
fc.cav <- decompose(ts.cav, type="additive")
fc.cor <- decompose(ts.cor, type="additive")
fc.rav <- decompose(ts.rav, type="additive")
# fc.hdr <- decompose(ts.hdr, type="additive")

# Add time series results to dataframes
df.cv <- df.cv %>% mutate(trend = as.numeric(fc.cav$trend),
                          seasonal = as.numeric(fc.cav$seasonal),
                          cor = as.numeric(fc.cav$x) - seasonal)
df.cr <- df.cr %>% mutate(trend = as.numeric(fc.cor$trend),
                          seasonal = as.numeric(fc.cor$seasonal),
                          cor = as.numeric(fc.cor$x) - seasonal)
df.rv <- df.rv %>% mutate(trend = as.numeric(fc.rav$trend),
                          seasonal = as.numeric(fc.rav$seasonal),
                          cor = as.numeric(fc.rav$x) - seasonal)
# df.hd <- df.hd %>% mutate(trend = as.numeric(fc.hdr$trend),
#                           seasonal = as.numeric(fc.hdr$seasonal),
#                           cor = as.numeric(fc.hdr$x) - seasonal)

# Merge it all together
df.mileage <- bind_rows(df.cv, df.cr, df.rv, df.hd) %>%
  mutate(label = paste0("Date: ", format.Date(date, "%Y-%m"), "<br>",
                        "Mileage: ", round(mileage, 1)," mpg"))

#
# Plot monthly gas mileage for each vehicle 
#
cols <- c("cav" = "#f8988f","cor"="#74c973", "rav"="#96aff2","hdr"="grey55")
p.mlg2 <- ggplot(df.mileage, aes(text=label)) + 
  geom_point(aes(x=date, y=mileage, group=car, color=car), size=1.2, alpha=0.5) + 
  geom_line(aes(x=date, y=trend, group=car, color=car), size=1.2, alpha=0.8) + 
  annotate("text",x=as.Date("2006-02-01"),y=34,label="Cavalier",
           color="#f8988f",size=5,alpha=0.9) +
  annotate("text",x=as.Date("2010-01-01"),y=37,label="Corolla",
         color="#74c973",size=5,alpha=0.9) +
  annotate("text",x=as.Date("2020-01-01"),y=18,label="Rav4",
         color="#96aff2",size=5,alpha=0.9) +
  annotate("text",x=as.Date("2023-03-01"),y=38,label="Highlander",
         color="grey55",size=5,alpha=0.9, hjust=1.0) +
  scale_color_manual(values = cols) +
  scale_y_continuous(limits = c(15,40)) +
  labs(x="Date",y="Gas Mileage (mpg)") +
  theme_minimal() +
  theme(legend.position="none") +
  theme(axis.text = element_text(size=10))

```

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
ggplotly(p.mlg2, tooltip="text")
```

<details>
<summary> <b> Details: Gas Mileage </b> </summary>
<br>
These histograms compare the monthly gas mileage aggregates to values quoted on https://www.fueleconomy.gov/ for each vehicle.

```{r, include=FALSE}

df.fuel <- data.frame(car = c("cav","cor","rav","hdr"),
                      city = c(21,27,22,36),
                      hwy = c(29,35,28,35),
                      cmb = c(24,30,25,36),
                      lab = c("Cavalier","Corolla","RAV4", "Highlander"))

p.mlg <- ggplot(df.cars) +
  geom_segment(data=df.fuel, aes(x=city, xend=city, y=0, yend=50, color=car)) +
  geom_text(data=df.fuel, aes(x=city, y=45, label="city",color=car), hjust=1.2,alpha=0.8) +
  geom_text(data=df.fuel, aes(x=hwy, y=45, label="hwy",color=car), hjust=1.2,alpha=0.8) +
  geom_text(data=df.fuel, aes(x=45, y=40, label=lab,color=car), hjust=1.0, size=5) +
  geom_segment(data=df.fuel, aes(x=hwy, xend=hwy, y=0, yend=50, color=car)) +
  geom_histogram( aes(x=mileage, fill=car), bins=40, alpha=0.8) +
  labs(x="Gas Mileage (mpg)", y="Count") +
  lims(x=c(15,45)) +
  scale_color_manual(values = cols) +
  scale_fill_manual(values = cols) +
  facet_wrap( ~car, ncol=1) +
  theme_minimal() +
  theme(legend.position="none", strip.text = element_blank())
  
```

```{r, echo=FALSE}
p.mlg
```


</details>

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE, message=FALSE, warning=FALSE}

# Time series object was created above
# Seasonal component to gas mileage
df.smileage <- df.mileage %>% 
  group_by(car, month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal))

p.smileage <- ggplot(df.smileage) + 
  geom_point(aes(x=month, y=seasonal, group=car, color=car), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=car, color=car), size=1, alpha=0.2) + 
  #scale_y_continuous(limits = c(15,40)) +
  labs(x="Month",y="Seasonal change in gas mileage (mpg)") +
  theme_minimal() +
  theme(legend.position="none",axis.text = element_text(size=10))

p.smileage2 <- ggplot(df.smileage) +
  geom_col(aes(x=month, y=seasonal, fill=seasonal >=0)) +
  geom_text(data=df.fuel, aes(x=11.9, y=2.5, label=lab), color="grey50", hjust=1.0, size=5) +
  facet_wrap(~car, ncol=1) +
  scale_fill_manual(values=pal.smiles) +
  labs(x="Month", y="Seasonal change in gas mileage (mpg)") +
  theme_minimal() +
  theme(axis.text = element_text(size=10), 
        legend.position="none",
        strip.text = element_blank())

diffcav <- max(df.smileage$seasonal[df.smileage$car == "cav"]) - 
           min(df.smileage$seasonal[df.smileage$car == "cav"])
diffcor <- max(df.smileage$seasonal[df.smileage$car == "cor"]) - 
           min(df.smileage$seasonal[df.smileage$car == "cor"])
diffrav <- max(df.smileage$seasonal[df.smileage$car == "rav"]) - 
           min(df.smileage$seasonal[df.smileage$car == "rav"])

```

<br>

Seasonal fluctuations in gas mileage were determined from a time series decomposition.  The values vary for each vehicle:
<ul>
<li>For the <b>Cavalier</b> the magnitude is  <b>`r format(diffcav,digits=1,nsmall=1)` mpg</b></li>
<li>For the <b>Corolla</b> the magnitude is <b>`r format(diffcor,digits=1,nsmall=1)` mpg</b></li>
<li>For the <b>RAV4</b> the magnitude is <b>`r format(diffrav,digits=1,nsmall=1)` mpg</b></li>
</ul>
```{r, echo=FALSE, message=FALSE, warning=FALSE}
p.smileage2
```
</details>

<br>
<br>
<br>

## Gas Prices


```{r, include=FALSE, eval=TRUE, warning=FALSE}
# Gas prices
df.rate <- df.miles

# Read file from eia.gov
df.gas <- read.csv(GAS_PRICES, skip = 5, header=FALSE) %>%
  mutate(date=as.POSIXct(paste0("1 ",V1), format="%d %b %Y")) %>%
  rename(mrate=V2) %>% 
  select(date,mrate) %>%
  filter(date >= min(df.rate$date))

# Combine
df.rate <- df.rate %>% left_join(df.gas) %>%
  mutate(label = paste0("Date: ", format.Date(date, "%Y-%m"), "<br>",
                        "AVG cost: $", round(mrate, 2),"<br>",
                        "AVG paid: $", round(rate, 2)))

#
# Plot gas prices over time
#
p.rate <- ggplot(df.rate, aes(text=label, group=1)) +
  geom_segment(aes(x=date, xend=date, y=rate, yend=mrate), size=0.3,color="grey75",alpha=0.7) +
  geom_line(aes(x=date, y=mrate), size=1.0, color="grey75", alpha=0.7) +
  geom_point(aes(x=date, y=rate), color="grey25", size=1.0, alpha=0.8) +
  #geom_line(aes(x=date, y=cor), color="indianred", size=1.2, alpha=0.8) +
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Date", y="Gas Price") +
  theme_minimal()

```

The plot below compares what we paid for gas with the national average monthly gas price for regular grade fuel according to the [US Energy Information Administration](https://www.eia.gov/opendata/v1/qb.php?category=240692&sdid=PET.EMM_EPMR_PTE_NUS_DPG.M).  Each data point represents the monthly aggregate for all gas purchased during a particular month, and the trendline indicates the national average.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
ggplotly(p.rate, tooltip = "text")
```

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE, message=FALSE, warning=FALSE}

# Create a timeseries object
ts.rate <- na.interp(ts(df.rate$rate, start = c(2005,4),frequency=12))
fc.rate <- decompose(ts.rate)
df.rate$trend <- as.numeric(fc.rate$trend)
df.rate$seasonal <- as.numeric(fc.rate$seasonal)
df.rate$cor <- as.numeric(fc.rate$x) - as.numeric(fc.rate$seasonal)

# 
df.rt <- df.rate %>% group_by(month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal, na.rm=TRUE)) %>%
  mutate(month = as.factor(month))

p.srate <- ggplot(df.rt) +
  geom_point(aes(x=month, y=seasonal), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=1), size=1, alpha=0.2) + 
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Month", y="Seasonal change in gas price") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))

p.srate2 <- ggplot(df.rt) +
  geom_col(aes(x=month, y=seasonal, fill=seasonal>=0)) +
  scale_y_continuous(labels=dollar_format()) +
  scale_fill_manual(values=pal.smiles) +
  labs(x="Month", y="Seasonal change in gas price") +
  theme_minimal() +
  theme(axis.text = element_text(size=10), 
        legend.position="none",
        strip.text = element_blank())
```

<br>
The seasonal fluctuations in the gas prices amount to <b> `r dollar(max(df.rt$seasonal)-min(df.rt$seasonal)) ` </b>, with the peak price in <b> `r month(which.max(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>, and the lowest in <b> `r month(which.min(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>. 

```{r, echo=FALSE, message=FALSE, warning=FALSE}
p.srate2
```
</details>

<br>
<br>
<hr>
<br>
<br>
<br>









