---
title: "Analysis of Gas Receipts"
author: "Nathan Hoteling"
date: "`r format(Sys.time(), '%B %Y')`"
output: html_document
---



```{r, include=FALSE}
p.miles <- ggplot(df.miles) +
  geom_line(aes(x=date, y=miles), color="grey20") +
  scale_y_continuous(limits=c(0,1.1*max(df.miles$miles)), labels=comma) +
  labs(x="Date", y="Monthly Miles") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

<br />

This analysis provides a summary of data collected from gas receipts for all vehicles I have owned from <b> `r format(min(df.bymonth$date, na.rm=TRUE), "%B %Y")` </b> to <b> `r format(max(df.bymonth$date, na.rm=TRUE), "%B %Y")`</b>, a span of <b> `r format(tlen, digits=1,nsmall=1)` </b> years.  Over this time period we drove a total of <b> `r format(tot_miles, digits=0,nsmal=0, scientific=FALSE, big.mark=",")` </b> miles and spent <b> `r dollar(tot_cost)` </b> on gas.

<br />





## Gas Prices


```{r, include=FALSE, eval=TRUE, warning=FALSE}
# Gas prices
df.rate <- df.miles

# Read file from eia.gov
df.gas <- read.csv("../data/U.S._All_Grades_All_Formulations_Retail_Gasoline_Prices_Monthly.csv",
                   skip = 5, header=FALSE) %>%
  mutate(date=my(V1)) %>% rename(mrate=V2) %>% select(date,mrate) %>%
  filter(date >= min(df.rate$date))

# Combine
df.rate <- df.rate %>% left_join(df.gas)

#
# Plot gas prices over time
#
p.rate <- ggplot(df.rate) +
  geom_segment(aes(x=date, xend=date, y=rate, yend=mrate), size=0.3,color="grey75",alpha=0.7) +
  geom_line(aes(x=date, y=mrate), size=1.2, color="grey75", alpha=0.7) +
  geom_point(aes(x=date, y=rate), color="grey25", size=1.0, alpha=0.8) +
  #geom_line(aes(x=date, y=cor), color="indianred", size=1.2, alpha=0.8) +
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Date", y="Gas Price") +
  theme_minimal()

```

The plot below compares what we paid for gas with the national average monthly gas price according to https://www.eia.gov/.  Each data point represents the monthly aggregate for all gas purchased during a particular month, and the trendline indicates the national average.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
p.rate
```

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE}

# Create a timeseries object
ts.rate <- na.interp(ts(df.rate$rate, start = c(2005,4),frequency=12))
fc.rate <- decompose(ts.rate)
df.rate$trend <- as.numeric(fc.rate$trend)
df.rate$seasonal <- as.numeric(fc.rate$seasonal)
df.rate$cor <- as.numeric(fc.rate$x) - as.numeric(fc.rate$seasonal)

# 
df.rt <- df.rate %>% group_by(month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal, na.rm=TRUE)) %>%
  mutate(month = as.factor(month))

p.srate <- ggplot(df.rt) +
  geom_point(aes(x=month, y=seasonal), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=1), size=1, alpha=0.2) + 
  scale_y_continuous(labels=dollar_format()) +
  labs(x="Month", y="Change in gas price") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

<br />
The seasonal fluctuations in the gas prices amount to <b> `r dollar(max(df.rt$seasonal)-min(df.rt$seasonal)) ` </b>, with the peak price in <b> `r month(which.max(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>, and the lowest in <b> `r month(which.min(df.rt$seasonal), label=TRUE, abbr=FALSE)` </b>. 

```{r, echo=FALSE}
p.srate
```
</details>

<br />


