---
title: "Total Miles"
---

```{r, include=FALSE}
p.miles <- ggplot(df.miles) +
  geom_line(aes(x=date, y=miles), color="grey20") +
  scale_y_continuous(limits=c(0,1.1*max(df.miles$miles)), labels=comma) +
  labs(x="Date", y="Monthly Miles") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

The plot below shows the number of miles driven per month for all vehicles over this period.  Note that gaps in the line chart represent months with no data; in some cases this was because we didn't purchase gas during that month and in others it is because gas receipts were lost.

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=8, fig.height=4}
p.miles
```

<details>
<summary> <b> Details: Seasonality </b> </summary>

```{r, include=FALSE}

# Create a timeseries object
ts.miles <- na.interp(ts(df.miles$miles, start = c(2005,4),frequency=12))
fc.miles <- decompose(ts.miles)
df.miles$trend <- as.numeric(fc.miles$trend)
df.miles$seasonal <- as.numeric(fc.miles$seasonal)
df.miles$cor <- as.numeric(fc.miles$x) - as.numeric(fc.miles$seasonal)

# 
df.smiles <- df.miles %>% group_by(month=month(date, label=TRUE, abbr=TRUE)) %>% 
  summarise(seasonal=mean(seasonal, na.rm=TRUE)) %>%
  mutate(month = as.factor(month))

p.smiles <- ggplot(df.smiles) +
  geom_point(aes(x=month, y=seasonal), size=4, alpha=0.8) +
  geom_line(aes(x=month, y=seasonal, group=1), size=1, alpha=0.2) + 
  labs(x="Month", y="Difference in miles driven") +
  theme_minimal() +
  theme(axis.text = element_text(size=10))
```

<br />
There are some seasonal patterns in the number of miles driven over the course of the year. 

```{r, echo=FALSE}
p.smiles
```
</details>

<br />